---
title: "Traffic Predictions - Time Series Analysis"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# 1. Data Overview

## 1.1 Loading Packages

```{r setup}
# Install necessary packages
library(fpp2)
library(zoo)
library(readr)
library(dplyr)
library(grid)
library(data.table)
library(stringr)
library(sjmisc)
library(TSstudio)
library(ggplot2)

set.seed(506)
```

## 1.2 Reading Data to Dataframe

```{r DataInput}
# Import Data - dateframe set to raw dataset from GitHub repository 

traffic<- read.csv("https://raw.githubusercontent.com/jvo024/traffic_predictions_ads506/main/traffic.csv")
traffic_df <- traffic 
head(traffic_df)
```

This data set contains 48,120 observations on vehicles passing through different junctions at different times and dates. The variables are as follows: DateTime, Junction, Vehicles, ID

# 2. Exploratory Data Analysis

## 2.1 Check for Null Values

```{r null}
print(which(is.na(traffic_df)))
```

No null values were identified from all 3 columns.

## 2.2 Summary of Data

```{r summary }
summary(traffic_df)
```

From an initial summary of the data set, we know that there are 4 junctions that were observed. We also can know that within an hour, up to 180 vehicles can be sighted.

## 2.3 Plot Time Series

Set the datatime (POSIXct) variable in the traffic dataframe

```{r Format DateTime}
traffic_df$DateTime <- as.POSIXct(traffic_df$DateTime, tz = "UTC")
```

Initial time series plot of first 250 observations on vehicles. We only use the first 250 observations because the data set is large. The frequency is set to 24 since there are 24 hours in each day.

```{r}
head <- head(traffic_df, 250)
vehicles.ts <- ts(head$Vehicles, frequency = 24)

autoplot(vehicles.ts)
```

## 2.4 Plot Time Series by Junction

To get some further information on the overall data set, the main data frame was subset into 4 based on the junction associated with the data and time. Data was aggregated into

```{r}
# Split by Junction
junc_1 <- traffic_df[traffic_df$Junction == 1, ]
junc_2 <- traffic_df[traffic_df$Junction == 2, ]

# Create Time series 
junc_1.ts <- ts(junc_1$Vehicles, frequency = 24)
junc_2.ts <- ts(junc_2$Vehicles, frequency = 24)

# Plot Junction Time Plots
ggarrange(
  
autoplot(junc_1.ts, main = "Junction 1") +
  labs(title = "Hourly Vehicles at Junction 1",
       x = "Time", 
       y = "# Vehicles "),

autoplot(junc_2.ts, main = "Junction 2")+ 
  labs(title = "Hourly Vehicles at Junction 2",
       x = "Time", 
       y = "# Vehicles "), 

nrow = 2, ncol = 1 
)

```

## 2.5 Plot autocorrelations of each Junction

```{r}

par(mfrow = c(2,1))
acf(junc_1.ts, main = "ACF - Junction 1")
acf(junc_2.ts, main = "ACF - Junction 2")

```

## 2.6 Monthly Vehicles at Each Junction

```{r}
## Compare vehicles per month at each junction 

# Junction 1
j1_11<- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(11)], ])$Vehicles)
j1_12 <- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(12)], ])$Vehicles)
j1_1 <- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(1)], ])$Vehicles)
j1_2 <- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(2)], ])$Vehicles)
j1_3 <- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(3)], ])$Vehicles)
j1_4 <- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(4)], ])$Vehicles)
j1_5 <- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(5)], ])$Vehicles)
j1_6 <- sum((junc_1[months(junc_1$DateTime) %in% month.name[c(6)], ])$Vehicles)

j1_vehicle_totals = c(j1_11, j1_12, j1_1, j1_2, j1_3, j1_4, j1_5, j1_6)
j1_months = c('Nov', 'Dec', 'Jan', 'Feb', 'Mar', 'April', 'May', 'June')

# Junction 2
j2_11<- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(11)], ])$Vehicles)
j2_12 <- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(12)], ])$Vehicles)
j2_1 <- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(1)], ])$Vehicles)
j2_2 <- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(2)], ])$Vehicles)
j2_3 <- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(3)], ])$Vehicles)
j2_4 <- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(4)], ])$Vehicles)
j2_5 <- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(5)], ])$Vehicles)
j2_6 <- sum((junc_2[months(junc_2$DateTime) %in% month.name[c(6)], ])$Vehicles)

j2_vehicle_totals = c(j2_11, j2_12, j2_1, j2_2, j2_3, j2_4, j2_5, j2_6)
j2_months = c('Nov', 'Dec', 'Jan', 'Feb', 'Mar', 'April', 'May', 'June')

# Plot bar chart of vehicles per month 
par(mfrow=c(1,2))
barplot(j1_vehicle_totals, 
        names.arg = j1_months,
        xlab = 'Month',
        ylab = 'Total Vehicles',
        main = 'Total Vehicles/Month at Junction 1',
        cex.names = 0.55)
barplot(j2_vehicle_totals, 
        names.arg = j2_months,
        xlab = 'Month',
        ylab = 'Total Vehicles',
        main = 'Total Vehicles/Month at Junction 2',
        cex.names = 0.55)
```

# 3. Pre-processing

## 3.1 Data Cleaning

The ID column is not correlated to the time series analysis that we will be conducting. Thus, we can opt to remove this variable while creating the model.

```{r Drop ID}
drop <- c("ID")
traffic_df = traffic_df[,!(names(traffic_df) %in% drop)]
```

## 3.2 Feature Engineering

```{r FeatureEngineeringSetUp}
# Create duplicate data set to perform feature engineering
df_fengin <- data.frame(traffic_df)
```

### 3.2.1 Aggregate Hourly to Daily

```{r Aggregate}

# Subset into Junctions from duplicate data frame 
j1 <- subset(df_fengin, Junction == 1)
j2 <- subset(df_fengin, Junction == 2)

# Aggregate Data to Daily
#j1
j1$Date <- as.Date(j1$DateTime)
j1<- aggregate(j1["Vehicles"], by=j1["Date"], sum)
j1[ , 'Junction'] <- 1

#j2
j2$Date <- as.Date(j2$DateTime)
j2<- aggregate(junc_2["Vehicles"], by=j2["Date"], sum)
j2[ , 'Junction'] <- 2

```

### 3.2.2 Identify Holidays - Create Dummy Variable

```{r Holidays }

# Create list of bank holiday values found via: https://www.calendardate.com/federal_holidays_2016.html
bank_holidays <- as.Date(c(

# 2015 
NewYears = "2015-01-01", 
MLK = "2015-01-19", 
Pres =  "2015-02-16",
Mem = "2015-05-25",
July4 = "2015-07-03", 
Labor = "2015-09-07", 
Columbus = "2015-10-12", 
Thank = "2015-11-26", 
XMas = "2016-12-25", #**
  
# 2016 
NewYears = "2016-01-01", 
MLK = "2016-01-18", 
Pres = "2016-02-15", 
Mem = "2016-05-30",
Labor = "2016-09-05", 
Columbus = "2016-10-10", 
Thank = "2016-11-24", 
XMas = "2016-12-26", 
  
# 2017
NewYears = "2015-01-01", 
NewYears = "2017-01-02",
MLK = "2017-01-16",
Pres =  "2017-02-20", 
Mem = "2017-05-29",
Labor = "2017-09-04",
Columbus = "2017-10-9",  
Thank = "2017-11-23", 
XMas = "2017-12-25" 
))

# Extract date from subsets
junc_1_date <- as.Date(junc_1$DateTime)
junc_2_date <- as.Date(junc_2$DateTime)

# Create Holiday column(in each subset) by testing if Holiday 
j1[ , 'Holiday'] <- as.Date(j1$Date) %in% bank_holidays
j2[ , 'Holiday'] <- as.Date(j2$Date) %in% bank_holidays

```


```{r}
## Initialize aggregated data as time series
# Weekly seasonality now that data is daily not hourly 
j1.ts <- ts(j1$Vehicles, frequency = 7)
j2.ts <- ts(j2$Vehicles, frequency = 7)
```

### 3.2.3 Identify Outliers

```{r}
# Check for Outliers using Boxplots

# Plot boxplots 
boxplot1 <- (ggplot(j1) +
               aes(x = "", y = Vehicles) +
               geom_boxplot(fill = "#0c4c8a") +
               ggtitle("Boxplot of Vehicles at Junction 1") +
               theme_minimal())

boxplot2 <-  (ggplot(j2) +
                aes(x = "", y = Vehicles) +
                geom_boxplot(fill = "#0c4c8a") +
                ggtitle("Boxplot of Vehicles at Junction 2") +
                theme_minimal())

ggarrange( boxplot1, boxplot2,  ncol = 2, nrow = 1)


# Extract outliers from boxplots 
outliers <- boxplot(j2$Vehicles,plot=FALSE)$out 
j1[ , 'outlier'] <- j1$Vehicles %in% outliers

# Test for dates that are a holiday and an outlier - none 
j1[(j1$outlier == TRUE & j1$Holiday== TRUE),]

# Which dates are associated with these outliers? 
print('Outlier Values of Vehicles occur on the following dates:')
print((j1[(j1$outlier == TRUE),])$Date)
```

```{r Time Plots}

# Examine Time Plots of aggregated data 
ggarrange(
  
## Junction 1  -  Shows positive additive trend and multiplicative seasonality
  autoplot(j1.ts) + 
    labs(title = "Vehicles at Junction 1",
         x = "Time", 
         y = "# Vehicles "),
  
## Junction 2  - Shows multiplicative trend? and seasonality
# Possible outlier around season 50
  autoplot(j2.ts) +
    labs(title = "Vehicles at Junction 2",
         x = "Time", 
         y = "# Vehicles "),
  nrow = 2, ncol = 1 
)

```

```{r ACF}

## Autocorrelation Plots of each Junction 
par(mfrow = c(2,1))
acf(j1.ts)   # ACF shows cyclical behavior, very high autocorrelation
acf(j2.ts)   # ACF shows definite cyclical behavior, very high autocorrelation 

```

# 4. Modeling

## 4.1 Partition Train and Validation Data

We split the data by junction to have separate train/test sets for each.
```{r Partition Data}

# 18 months of data for training, 1 month testing

## Junction 1
j1.split <- ts_split(ts.obj = j1.ts, sample.out = 122)
j1.train <- j1.split$train
j1.test <- j1.split$test

## Junction 2
j2.split <- ts_split(ts.obj = j2.ts, sample.out = 122)
j2.train <- j2.split$train
j2.test <- j2.split$test
```

## 4.2 Naive Model (Baseline) 

```{r Naive}
# Forecast horizon for each is one month ahead 
j1.naive <- naive(j1.train, h = 122)
#summary(j1.naive)

j2.naive <- naive(j2.train, h = 122)
#summary(j2.naive)

# accuracy
j1.naive.accuracy <- accuracy(j1.naive, j1.test)
j2.naive.accuracy <- accuracy(j2.naive, j2.test)

# Calculate Train RMSE Values
j1.naive.train.RMSE <- j1.naive.accuracy[3]
j2.naive.train.RMSE <- j2.naive.accuracy[3]

# Calculate Test RMSE Values
j1.naive.test.RMSE <- j1.naive.accuracy[4]
j2.naive.test.RMSE <- j2.naive.accuracy[4]

```

## 4.3 Seasonal Naive 
```{r Seasonal Naive}

j1.snaive <- snaive(j1.train, h = 122)
#summary(j1.naive)

j2.snaive <- snaive(j2.train, h = 122)
#summary(j2.naive)

# accuracy
j1.snaive.accuracy <- accuracy(j1.snaive, j1.test)
j2.snaive.accuracy <- accuracy(j2.snaive, j2.test)

# Calculate Train RMSE Values
j1.snaive.train.RMSE <- j1.snaive.accuracy[3]
j2.snaive.train.RMSE <- j2.snaive.accuracy[3]

# Calculate Test RMSE Values
j1.snaive.test.RMSE <- j1.snaive.accuracy[4]
j2.snaive.test.RMSE <- j2.snaive.accuracy[4]
  
```

## 4.4 Smoothing Method

```{r}
# Smoothing with R-specified error-trend-seasonality combination 

j1.sm <- ets(j1.train, model = "ZZZ")
j1.sm.pred <- forecast(j1.sm, h = 122)
#summary(j1.se)

j2.sm <- ets(j2.train, model = "ZZZ")
j2.sm.pred <- forecast(j2.sm, h = 122)
#summary(j2.se)

# accuracy
j1.sm.accuracy <- accuracy(j1.sm.pred, j1.test)
j2.sm.accuracy <- accuracy(j2.sm.pred, j2.test)

# Calculate Train RMSE Values
j1.sm.train.RMSE <- j1.sm.accuracy[3]
j2.sm.train.RMSE <- j2.sm.accuracy[3]

# Calculate Test RMSE Values
j1.sm.test.RMSE <- j1.sm.accuracy[4]
j2.sm.test.RMSE <- j2.sm.accuracy[4]

```

## 4.5 Holt's Trend Method

```{r}

j1.holt <- holt(j1.train, h = 122)
#summary(j1.holt)

j2.holt <- holt(j2.train, h = 122)
#summary(j2.holt)

# accuracy
j1.holt.accuracy <- accuracy(j1.holt, j1.test)
j2.holt.accuracy <- accuracy(j2.holt, j2.test)

# Calculate Train RMSE Values
j1.holt.train.RMSE <- j1.holt.accuracy[3]
j2.holt.train.RMSE <- j2.holt.accuracy[3]

# Calculate Test RMSE Values
j1.holt.test.RMSE <- j1.holt.accuracy[4]
j2.holt.test.RMSE <- j2.holt.accuracy[4]

```

## 4.5 Regression Method (with trend and seasonality)

```{r}

j1.regression <- tslm(j1.train ~ trend + season)
#summary(j1.regression)

j2.regression <- tslm(j2.train ~ trend + season)
#summary(j2.regression)

j1.regression.predictions <- forecast(j1.regression, h= 122)
j2.regression.predictions <- forecast(j2.regression, h= 122)

# accuracy

# Calculate Train RMSE Values
j1.regression.train.RMSE <- sqrt(mean(j1.regression$residuals^2))
j2.regression.train.RMSE <- sqrt(mean(j2.regression$residuals^2))

# Calculate Test RMSE Values
j1.regression.test.RMSE <- accuracy(j1.regression.predictions, j1.test)[4]
j2.regression.test.RMSE <- accuracy(j2.regression.predictions, j2.test)[4]
```

## 4.6 Auto ARIMA 
```{r AutoARIMA}

j1.autoarima <- auto.arima(j1.train)
#summary(j2.autoarima)

j2.autoarima <- auto.arima(j2.train)
#summary(j2.autoarima)

j1.autoarima.predictions <- forecast(j1.autoarima, h= 122)
j2.autoarima.predictions <- forecast(j2.autoarima, h= 122)

# accuracy

# Calculate Train RMSE Values
j1.autoarima.train.RMSE <- sqrt(mean(j1.autoarima$residuals^2))
j2.autoarima.train.RMSE <- sqrt(mean(j2.autoarima$residuals^2))

# Calculate Test RMSE Values
j1.autoarima.test.RMSE <- accuracy(j1.autoarima.predictions, j1.test)[4]
j2.autoarima.test.RMSE <- accuracy(j2.autoarima.predictions, j2.test)[4]
```

## 4.7 Auto ARIMA with External Variables

```{r AutoARIMA with predictors}
#Change logical dummy variables to binary 

j1$outlier<- as.integer(as.logical(j1$outlier))
j1$Holiday<- as.integer(as.logical(j1$Holiday))
j2$Holiday<- as.integer(as.logical(j2$Holiday))

## Partition Data with External Variables
# Junction 1 predictors include holidays and outliers
# Junction 2 only holidays since there are no outliers
j1.train.external <- j1[1:486,] %>%
  select(-Date, -Junction)

j2.train.external <- j2[1:486,]%>%
  select(-Date, -Junction)

## Set Outcomes
j1.outcome <- j1.train.external$Vehicles
j2.outcome <- j2.train.external$Vehicles

## Set Predictors 
j1.predictors <- as.matrix(j1.train.external[, 2:3])
j2.predictors <- as.matrix(j2.train.external[, 2])

## Create Models 
j1.autoarima.ex <- Arima(j1.outcome, xreg = (j1.predictors))
#summary(j1.autoarima.ex)

j2.autoarima.ex <- auto.arima(j2.outcome, xreg = as.numeric(j2.predictors))
#summary(j2.autoarima.ex)


#j1.autoarima.ex.predictions <- forecast(j1.autoarima.ex, h= 122)
#j2.autoarima.ex.predictions <- forecast(j2.autoarima.ex, h= 122)

j1.test.external <- j1[487:608,] %>%
  select(-Date, -Junction)

j2.test.external <- j2[487:608,]%>%
  select(-Date, -Junction)

j1.predictors <- as.matrix(j1.test.external[, 2:3])
j2.predictors <- as.matrix(j2.test.external[, 2])

j1.autoarima.ex.pred <- forecast(j1.autoarima.ex, h = 122, xreg = (j1.predictors))
j2.autoarima.ex.pred <- forecast(j2.autoarima.ex, h = 122, xreg = (j2.predictors))

# accuracy

# Calculate Train RMSE Values
j1.autoarima.ex.train.RMSE <- sqrt(mean(j1.autoarima.ex$residuals^2))
j2.autoarima.ex.train.RMSE <- sqrt(mean(j2.autoarima.ex$residuals^2))

# Calculate Test RMSE Values
j1.autoarima.ex.test.RMSE <- sqrt(mean(j1.autoarima.ex.pred$residuals^2))
j2.autoarima.ex.test.RMSE <- sqrt(mean(j2.autoarima.ex.pred$residuals^2))


```

## 4.8 Neural Network 

```{r NNet}

j1.nnetar <- nnetar(j1.train, P = 11)
j2.nnetar <- nnetar(j2.train, P = 11)

j1.nnetar.pred <- forecast(j1.nnetar, h = 122)
j2.nnetar.pred <- forecast(j2.nnetar, h = 122)

# accuracy

j1.holt.accuracy <- accuracy(j1.nnetar.pred, j1.test)
j2.holt.accuracy <- accuracy(j2.nnetar.pred, j2.test)

# Calculate Train RMSE Values
j1.nnetar.train.RMSE <- sqrt(mean(j1.nnetar$residuals^2))
j2.nnetar.train.RMSE <- sqrt(mean(j2.nnetar$residuals^2))

# Calculate Test RMSE Values
j1.nnetar.test.RMSE <- j1.holt.accuracy[4]
j2.nnetar.test.RMSE <- j2.holt.accuracy[4]


```

# 5. Model Results

```{r}

j1.train.results <- data.frame(
  Model = c("Naive", "Seasonal Naive", "Exponential Smoothing", "Holt Winters",
            "Regression", "Auto ARIMA", "Auto ARIMA with predictors", "Neural Net"),
  RMSE.Junction1 = c(j1.naive.train.RMSE, j1.snaive.train.RMSE, j1.sm.train.RMSE, j1.holt.train.RMSE, j1.regression.train.RMSE,
           j1.autoarima.train.RMSE, j1.autoarima.ex.train.RMSE, j1.nnetar.train.RMSE)
  )

j2.train.results <- data.frame(
  Model = c("Naive", "Seasonal Naive", "Exponential Smoothing", "Holt Winters",
            "Regression", "Auto ARIMA", "Auto ARIMA with predictors", "Neural Net"),
  RMSE.Junction2 = c(j2.naive.train.RMSE, j2.snaive.train.RMSE, j2.sm.train.RMSE, j2.holt.train.RMSE, j2.regression.train.RMSE,
           j2.autoarima.train.RMSE, j2.autoarima.ex.train.RMSE, j2.nnetar.train.RMSE)
  )

j1.test.results <- data.frame(
  Model = c("Naive", "Seasonal Naive", "Exponential Smoothing", "Holt Winters",
            "Regression", "Auto ARIMA", "Auto ARIMA with predictors", "Neural Net"),
  RMSE.Junction1 = c(j1.naive.test.RMSE, j1.snaive.test.RMSE, j1.sm.test.RMSE, j1.holt.test.RMSE, j1.regression.test.RMSE,
           j1.autoarima.test.RMSE, j1.autoarima.ex.test.RMSE, j1.nnetar.test.RMSE)
  )

j2.test.results <- data.frame(
  Model = c("Naive", "Seasonal Naive", "Exponential Smoothing", "Holt Winters",
            "Regression", "Auto ARIMA", "Auto ARIMA with predictors", "Neural Net"),
  RMSE.Junction2 = c(j2.naive.test.RMSE, j2.snaive.test.RMSE, j2.sm.test.RMSE, j2.holt.test.RMSE, j2.regression.test.RMSE,
           j2.autoarima.test.RMSE, j2.autoarima.ex.test.RMSE, j2.nnetar.test.RMSE)
  )

# Junction 1 Results
j1.train.results
j1.test.results

# Junction 2 Results
j2.train.results
j2.test.results
```

```{r}
# Auto ARIMA and Smoothing outperformed for both junctions 

# Plot best two models for Junction 1 
autoplot(j1.train) +
  autolayer(j1.sm.pred, color = "green", series = "Smoothing", alpha = 0.5) +
  autolayer(forecast(j1.autoarima, h = 30), color = "blue", series = "ARIMA", alpha = 0.5) + 
  autolayer(j1.test, color = "red", series = "Actual", alpha = 0.5) +
  coord_cartesian(xlim = c(80, 88 )) 

# Plot best two models for Junction 1 
autoplot(j2.train) +
  autolayer(j2.sm.pred, color = "green", series = "Smoothing", alpha = 0.5) +
  autolayer(forecast(j2.autoarima, h = 30), color = "blue", series = "ARIMA", alpha = 0.5) + 
  autolayer(forecast(j2.regression, h = 30), color = "yellow", series = "Regression", alpha = 0.5) +
  autolayer(j2.test, color = "red", series = "Actual", alpha = 0.5) +
  coord_cartesian(xlim = c(80, 88 )) 
```
